\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{amssymb}
\usepackage{parskip}
\emergencystretch=3em
\sloppy

\definecolor{cmdbg}{gray}{0.93}
\newcommand{\cmd}[1]{\colorbox{cmdbg}{\texttt{#1}}}
\newcommand{\field}[1]{\texttt{#1}}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small Prompt Provenance Guide}
\fancyhead[R]{\small MSAI-LLM-PERFORMANCE}
\fancyfoot[C]{\small Page \thepage}

\title{Reconstructing the Exact LLM Inputs\\
\large How to read the benchmark questions from the repository's code and metadata\\
\normalsize Repository: \texttt{github.com/ericmedlock/MSAI-LLM-PERFORMANCE} \quad Guide version: 2026-07-11}
\author{ITCS 6881 --- LLM Execution Architecture Benchmark}
\date{}

\begin{document}
\maketitle
\thispagestyle{fancy}

\section*{The one-paragraph answer}

Every question the model ever sees is stored in plain text in the repository, in
three composable layers: (1) the \textbf{task manifests} under \texttt{tasks/}
hold each question's exact user-message text in the \field{prompt} field; (2) the
\textbf{frozen system prompts} under \texttt{prompts/} hold the per-role
instructions that precede it; (3) \texttt{config/config.yaml} pins the decoding
parameters (temperature, seed, context, output budget). On top of that, every
recorded run in \texttt{results/*.jsonl} carries a \field{raw\_trace} of the
messages \emph{actually sent}, so any historical run can be audited without
reconstruction. Nothing is generated at run time; reproduction is a matter of
reading files, not re-running code.

\section{Layer 1 --- the task manifests (\texttt{tasks/})}

\begin{tabular}{p{2.1in}p{3.9in}}
\textbf{File} & \textbf{Contents}\\
\hline\\[-8pt]
\texttt{manifest.json} & the frozen 15-item \emph{baseline} suite (5 GSM8K / 5 HumanEval / 5 HotpotQA)\\
\texttt{frontier\_v2\_manifest.json} & the frozen 36-item \emph{frontier} tier (12 AIME 2025 / 12 BigCodeBench-Hard / 12 MuSiQue open-book)\\
\texttt{frontier\_manifest.json}, \texttt{frontier\_external\_manifest.json} & superseded candidate generations, kept for provenance\\
\end{tabular}

Each entry in a manifest's \field{tasks} array has these fields:

\begin{itemize}
  \item \field{task\_id} --- the stable key every result row references (e.g.\ \texttt{fx2-mathA-001}).
  \item \field{prompt} --- \textbf{the complete user-message text, verbatim.} What
    you read here is byte-for-byte what the model receives as the user turn ---
    including the answer-format instruction line, and (for multihop items) the
    embedded context passages, and (for code items) the function signature and
    docstring to complete.
  \item \field{answer} --- the gold answer for exact/normalized matching
    (\texttt{null} for code items, which grade by unit test instead).
  \item \field{grading} --- for code items: \field{entry\_point} (the function
    name) and \field{test} (the full test source; the grader executes
    \emph{model code} + \emph{test} + \texttt{check(entry\_point)} in a
    subprocess).
  \item \field{source\_id} --- upstream identity, e.g.\
    \texttt{aime25:0}, \texttt{bigcodebench-hard-v0.1.4:BigCodeBench/16},
    \texttt{musique:2hop\_\_123\ldots\ (with context)}, \texttt{gsm8k/main train[0]}.
  \item \field{id\_status} --- provenance verdict
    (\texttt{FROZEN\_CALIBRATED\_IN\_BAND}, \texttt{VERIFIED\_UPSTREAM\_EXACT} /
    \texttt{\_ADAPTED}, \texttt{PROJECT\_AUTHORED}) --- see
    \texttt{scripts/verify\_task\_ids.py} and the Amendment Log.
  \item \field{selection\_reason} --- why the item is in the suite (human-readable
    audit trail, including calibration history).
  \item \field{reference\_solution} --- (code items) the upstream canonical
    solution, used only to self-verify the test at build time; never shown to the model.
\end{itemize}

\section{Layer 2 --- the frozen system prompts (\texttt{prompts/})}

Each architecture role has its own version-controlled system prompt:

\begin{tabular}{p{3.15in}p{2.85in}}
\texttt{prompts/monolithic/system.txt} & the single-shot solver\\
\texttt{prompts/agentic/executor\_system.txt} & the agentic executor\\
\texttt{prompts/agentic/verifier\_system.txt} & the agentic verifier\\
\texttt{prompts/swarm/peer\_system.txt} & every swarm peer (identical for all three)\\
\texttt{prompts/judge/system.txt} & the LLM-as-judge (post-processing only)\\
\end{tabular}

These files are frozen; \texttt{prompts/README.md} documents the intentional
role differences.

\section{Layer 3 --- message composition per architecture}

The harness composes exactly two-role conversations (one system + one user
message per call). From the backend sources (\texttt{backends/*.py}):

\textbf{Monolithic} --- one call:
\begin{quote}\ttfamily\small
system = prompts/monolithic/system.txt\\
user \ \ = manifest prompt (verbatim)
\end{quote}

\textbf{Agentic} (Executor + Verifier, max 2 loops) --- first executor call uses the
manifest prompt verbatim; the verifier then receives:
\begin{quote}\ttfamily\small
Task:\{question\}\textbackslash n\textbackslash nCandidate answer:\textbackslash n\{candidate\}
\end{quote}
If the verifier does not reply starting with \texttt{APPROVE}, the executor is
re-invoked with the templated revision message:
\begin{quote}\ttfamily\small
Your previous answer was:\textbackslash n\{candidate\}\textbackslash n\textbackslash nVerifier feedback: \{feedback\}\textbackslash nProvide a corrected final answer.
\end{quote}

\textbf{Swarm} --- three parallel peers, each: peer system prompt + manifest prompt
verbatim; the only difference between peers is the sampling seed
(\texttt{seed = base\_seed + peer\_index}); answers are majority-voted.

\textbf{Judge} (secondary metric) --- composed by \texttt{harness/judge.py} from the
judge system prompt + the task, gold answer, and the recorded model answer.

\textbf{Decoding pins (all calls):} temperature 0.0, seed 42 (+peer offset),
\texttt{num\_ctx} 8192, \texttt{max\_tokens} 6144 --- from
\texttt{config/config.yaml}, whose raw bytes are hashed into every row's
\field{config\_hash}.

\section{Recipes}

\textbf{Print one task's exact user message} (repo root, any machine):
\begin{quote}\ttfamily\small
./.venv/bin/python -c "import json; \\
\hspace*{1em}m=json.load(open('tasks/frontier\_v2\_manifest.json')); \\
\hspace*{1em}print([t for t in m['tasks'] if t['task\_id']=='fx2-mathA-001'][0]['prompt'])"
\end{quote}

\textbf{Dump every question to a file} (e.g.\ for a report appendix):
\begin{quote}\ttfamily\small
./.venv/bin/python -c "import json; \\
\hspace*{1em}m=json.load(open('tasks/frontier\_v2\_manifest.json')); \\
\hspace*{1em}print('\textbackslash n\textbackslash n'.join(f\textquotesingle=== \{t[\textquotesingle task\_id\textquotesingle]\} (\{t[\textquotesingle source\_id\textquotesingle]\}) ===\textbackslash n\{t[\textquotesingle prompt\textquotesingle]\}\textquotesingle\ for t in m['tasks']))" \\
\hspace*{1em}> all\_frontier\_questions.txt
\end{quote}

\textbf{Audit what a historical run actually sent} --- join a result row back to its
inputs via \field{task\_id}, or read the row's own \field{raw\_trace}, which
records each agent call's user message and output:
\begin{quote}\ttfamily\small
./.venv/bin/python -c "import json; \\
\hspace*{1em}r=[json.loads(l) for l in open('results/frontier-v2-local-ollama-14b.jsonl')]; \\
\hspace*{1em}row=[x for x in r if x['task\_id']=='fx2-mathA-001' and x['backend']=='agentic'][0]; \\
\hspace*{1em}print(json.dumps(row['raw\_trace'], indent=2))"
\end{quote}

\textbf{Byte-exact historical reproduction:} every row carries
\field{config\_hash}; check out the commit where \texttt{config/config.yaml}
hashes to that value (v1 freeze: tag \texttt{data-freeze-2026-07-01}) and the
three layers above are guaranteed identical to what ran.

\section{Worked example --- \texttt{fx2-mathA-001}}

Manifest fields (from \texttt{tasks/frontier\_v2\_manifest.json}):

\begin{quote}\ttfamily\small
task\_id:\ \ fx2-mathA-001\\
domain:\ \ \ math \quad tier: frontier\\
source\_id: aime25:0 \quad (AIME 2025, problem index 0 in math-ai/aime25)\\
id\_status: FROZEN\_CALIBRATED\_IN\_BAND\\
answer:\ \ \ 70
\end{quote}

The \field{prompt} field --- and therefore the exact user message, shown here
verbatim including its LaTeX-style math notation, which is passed through as
plain text:

\begin{quote}\ttfamily\small
Find the sum of all integer bases \$b>9\$ for which \$17\_b\$ is a divisor of \$97\_b.\$\\[4pt]
Give the final answer as a single integer on the last line.
\end{quote}

For a \emph{monolithic} run, the model receives
\texttt{prompts/monolithic/system.txt} as the system message and the text above
as the user message, decoded at temperature 0.0, seed 42, output cap 6144 ---
nothing else. For \emph{swarm}, three peers receive the identical pair with seeds
42/43/44. For \emph{agentic}, the executor starts from the same pair and the
verifier/revision templates of \S3 take over from there --- with the exact
realized sequence stored in the run row's \field{raw\_trace}.

\section*{Upstream provenance chain}

\texttt{source\_id} $\rightarrow$ fetch/verify tooling: baseline items are
verified against their public sources by \texttt{scripts/verify\_task\_ids.py};
frontier items were built (and self-graded through the real grader) by
\texttt{scripts/build\_frontier\_v2.py} from HuggingFace datasets
(\texttt{math-ai/aime25}, \texttt{bigcode/bigcodebench-hard},
\texttt{dgslibisey/MuSiQue}). The full re-sourcing history and every deviation is
in \texttt{PRE\_REGISTRATION.md}'s Amendment Log.

\end{document}
