\documentclass[sigconf,nonacm]{acmart}

\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{booktabs}
\emergencystretch=3em
\hbadness=10000
\sloppy

\begin{document}

\title{Calibrating a Frontier Task Tier for Architecture-Sensitive LLM Benchmarking}
\subtitle{Status Report II --- baseline results, full-size probe, and the frozen frontier-v2 tier}

\author{Eric Medlock}
\affiliation{%
  \institution{University of North Carolina at Charlotte}
  \department{ITCS 6881 Independent Study}
  \city{Charlotte}
  \state{North Carolina}
  \country{USA}
}
\email{emedlock@charlotte.edu}

\begin{abstract}
The pre-registered study compares monolithic, agentic (Executor+Verifier), and swarm
(3-peer majority vote) LLM execution architectures under pinned model, prompts, and
decoding. The v1 baseline (225 runs, frozen 2026-07-01) saturated: the pinned 14B model
solves the classic benchmark items nearly perfectly, leaving architectures nothing to
differentiate on. This report documents the construction and calibration of a
\emph{frontier tier} --- tasks selected so that monolithic single-pass accuracy lands in
the 0.4--0.7 band where iteration and voting can matter. A 330-run full-size probe
(DeepSeek-R1-distill 32B) exposed both failure directions: MBPP code items saturated
(100\%) while closed-book MuSiQue floored (6\%), and no architecture rescued the floor.
A three-round calibration campaign (350 monolithic runs at the pinned 14B) re-sourced
math to AIME~2025 (42\%), code to BigCodeBench-Hard (48\%), and multihop to MuSiQue
\emph{with} supporting passages (67\%); the resulting 36-item tier is frozen. Two
methodological findings fell out: per-item accuracy is near-binary under temperature-0
decoding (forcing the calibration band to the domain aggregate), and 47\% of frontier
failures are output-budget exhaustion rather than wrong reasoning. Two empirical
surprises: the 14B distill \emph{outperforms} its 32B sibling on the same MATH-500 slice
(88\% vs.\ 65\%), and on floored knowledge tasks agentic/swarm wrappers scored
\emph{zero} where monolithic scraped 17\%. The confirmatory tier$\times$architecture
sweep now moves to university HPC.
\end{abstract}

\maketitle

\section{Context}
Design, hypotheses (task-dependent superiority, ``candidate~D''), conditions, and
metrics are frozen in \texttt{PRE\_REGISTRATION.md} (2026-07-01 07:31 EDT); every
deviation since is in its Amendment Log. Architecture is the only variable: one pinned
model (DeepSeek-R1 14B distill, Q4\_K\_M, digest recorded), frozen prompts, temperature
0.0, fixed seed, $n_{ctx}=8192$, max output 6144 tokens, $N=5$ trials per cell. The
harness grades binarily and automatically (exact numeric / unit tests / normalized
string), writes one crash-safe JSONL row per run, and is idempotent.

\section{v1 baseline (frozen 2026-07-01)}
225 runs: 15 items (5 GSM8K / 5 HumanEval / 5 HotpotQA) $\times$ 3 architectures
$\times$ $N{=}5$, local cell (Apple M5 Max, LM Studio/Metal).

\begin{table}[h]
\caption{v1 baseline, all 15 items (mean$\pm$std across runs).}
\label{tab:v1}
\begin{tabular}{lrrrr}
\toprule
Architecture & Acc. & Latency (s) & Tokens & tok/s\\
\midrule
Monolithic & 93\% & $15.2\pm17.5$ & $832\pm850$ & 44.9\\
Agentic    & \textbf{100\%} & $32.5\pm32.0$ & $1961\pm1662$ & 44.5\\
Swarm      & 93\% & $32.9\pm33.1$ & $2649\pm2247$ & 63.0\\
\bottomrule
\end{tabular}
\end{table}

The agentic verifier fixed every monolithic error (notably the one item all
single-pass runs missed), at $\sim$2.4$\times$ token cost; the swarm's parallel peers
raised throughput (63 tok/s) but its majority vote failed exactly where peers agreed on
the same wrong answer. The problem is the 93--100\% ceiling itself: with at most one
hard item, the design cannot observe the \emph{task-dependent} effects it
pre-registered. Hence the frontier tier: a second, harder task class kept objectively
auto-graded, selected by a pre-declared empirical rule --- monolithic single-pass
accuracy in $[0.4, 0.7]$ --- the band where self-correction and voting have room to act.

\section{Full-size probe: three regimes}
Before calibrating at the pinned 14B, the 22 externally-sourced candidates
(MATH-500 L$\geq$3 / MBPP-sanitized / closed-book MuSiQue) were probed at full size:
DeepSeek-R1-distill-\textbf{32B}, all three architectures, $N{=}5$ --- 330 runs
($\sim$14\,h). The probe exposed both failure directions at once
(Table~\ref{tab:probe}).

\begin{table}[h]
\caption{32B probe: accuracy by domain $\times$ architecture.}
\label{tab:probe}
\begin{tabular}{lrrrl}
\toprule
Domain & Mono & Agentic & Swarm & Verdict\\
\midrule
Math (MATH-500)        & 70\% & 63\% & 63\% & in band\\
Code (MBPP)            & 100\% & 100\% & 100\% & \textbf{ceiling}\\
Multihop (closed-book) & 17\% & 0\% & 0\% & \textbf{floor}\\
\bottomrule
\end{tabular}
\end{table}

Two observations survive any later analysis. First, \emph{architecture cannot
substitute for knowledge}: closed-book MuSiQue asks for facts the model does not hold,
and iterating or voting over ignorance produced strictly \emph{worse} results than a
single pass (0/60 vs.\ 5/30) --- the verifier ``corrects'' toward confident
hallucinations, and peers vote on shared ones. Second, on the one in-band domain
(math), monolithic \emph{beat} both wrappers at 32B (70\% vs.\ 63\%) --- an early,
exploratory contradiction of the hypothesis's direction that the confirmatory 14B sweep
will test properly. Mean cost ordering matched pre-registration sub-prediction~1:
1610 / 2706 / 2884 tokens per run (mono/agentic/swarm).

\section{Calibration campaign and freeze}
Calibration ran monolithic-only, $N{=}5$, against the \emph{pinned 14B} (the model
under test). Three rounds, 350 runs total (Table~\ref{tab:calib}).

\begin{table}[h]
\caption{Domain aggregates per calibration round (pinned 14B). Bold = in band, frozen.}
\label{tab:calib}
\begin{tabular}{llrl}
\toprule
Round & Domain source & Agg. & Action\\
\midrule
1 & MATH-500 L3--5 mix        & 88\% & re-source harder\\
1 & BigCodeBench (stdlib)     & 89\% & re-source harder\\
1 & MuSiQue + context         & \textbf{67\%} & \textbf{freeze}\\
2 & MATH-500 L5-only          & 80\% & re-source harder\\
2 & BigCodeBench-\textbf{Hard} & \textbf{48\%} & \textbf{freeze}\\
3 & \textbf{AIME 2025}        & \textbf{42\%} & \textbf{freeze}\\
\bottomrule
\end{tabular}
\end{table}

Two amendments shaped the rules (both logged before the affected runs):

\textbf{Per-item binarity $\rightarrow$ domain-aggregate band.} Under temperature 0.0
with a fixed seed, the five trials of an item are near-identical: observed per-item
accuracies were 5/5 or 0/5 in 66 of 68 item-round combinations. A per-item 0.4--0.7
band is therefore unsatisfiable as written; the band was restated at the domain
aggregate, with whole-domain re-sourcing from pre-declared harder tiers
(MATH-500 $\rightarrow$ L5 $\rightarrow$ AIME~2025; BCB $\rightarrow$ BCB-Hard) and an
explicit prohibition on item-level outcome selection.

\textbf{Provided context for multihop.} MuSiQue is calibrated upstream as reading
comprehension; serving it closed-book turns it into a parametric-recall lottery (the
32B floor). v2 supplies the supporting passages plus distractors within a fixed
context budget, making the domain reasoning-bound. It calibrated to 67\% immediately.

The frozen tier (\texttt{tasks/frontier\_v2\_manifest.json}, 2026-07-09): 12 AIME~2025
math items (42\%), 12 BigCodeBench-Hard code items (48\%), 12 MuSiQue-with-context
multihop items (67\%); tier aggregate 52\%. Every domain contains genuinely solved and
genuinely unsolved items --- the property the tier$\times$architecture analysis needs.
AIME~2025 additionally post-dates the pinned model's training, removing the
data-contamination concern the MATH-500 items carry.

\section{Failure taxonomy at the frontier}
Of the 86 incorrect rows in the frozen tier's calibration, 46 are reasoning errors but
\textbf{40 (47\%) are output-budget exhaustions}: the model consumed all 6144 output
tokens on chain-of-thought and was truncated before emitting any final answer. (The
same mechanism forced the 2048$\rightarrow$6144 amendment during the v1 pilot; at
frontier difficulty it recurs at 6144.) This matters for interpretation: part of what
``too hard for the pinned model'' \emph{means} under a fixed budget is
``cannot finish reasoning in time.'' It is also an architecture-relevant mechanism ---
whether a verifier retry or a differently-seeded peer can land an answer inside the
budget where a single pass cannot is precisely a tier$\times$architecture question.

\section{Findings so far}
\begin{enumerate}
\item \textbf{Baseline saturation is real and model-size-robust}: classic items
(GSM8K/HumanEval/MBPP-class) are ceiling-level for both 14B and 32B; architecture
comparisons on such suites measure noise.
\item \textbf{Capability is not monotonic in scale per-domain}: the 14B distill scored
88\% on the same MATH-500 slice where the 32B scored 65\%, while the 32B was far
stronger on closed-book recall (its 6\% floor still beat the 14B's expected zero).
Difficulty bands are model-relative; tiers must be calibrated per model.
\item \textbf{Architecture cannot fix missing knowledge}, and can amplify its absence
(0\% agentic/swarm vs.\ 17\% mono on the floored domain).
\item \textbf{Budget exhaustion is a first-class failure mode} at frontier difficulty
(47\% of failures), distinct from wrong reasoning and reportable separately.
\item (Exploratory) \textbf{monolithic led both wrappers on in-band math at 32B};
whether this holds at the pinned 14B is exactly what the confirmatory sweep tests.
\end{enumerate}

\section{Threats to validity}
Single model family (DeepSeek-R1 distills) and single environment so far (Metal/LM
Studio; the pre-registered cloud cell is pending). Domain-aggregate calibration keeps
selection mechanical but leaves per-domain aggregates at different band positions (42
/ 48 / 67\%), so cross-domain accuracy comparisons remain descriptive. AIME~2025
avoids contamination, but BigCodeBench-Hard and MuSiQue pre-date the model's cutoff.
$N{=}5$ under near-deterministic decoding characterizes serving jitter, not sampling
variance. The 32B architecture observations use uncalibrated v1-candidate items and
are exploratory only.

\section{Next steps}
(1)~Confirmatory sweep on university HPC: frozen tier $\times$ 3 architectures
$\times$ $N{=}5$ at the pinned 14B (monolithic rows seeded from calibration), per
\texttt{docs/HPC\_RUNBOOK.md}; (2)~exploratory scaling legs (32B, Qwen3.6-35B-A3B) at
$N{=}2$; (3)~LLM-as-judge secondary pass (Llama-3.2-3B); (4)~the pre-registered Azure
cloud cell; (5)~future work: a tool-equipped v2 study arm where GAIA-style tasks
become applicable --- rejected for the current tier because it benchmarks tool
orchestration, which the no-tool design deliberately excludes.

\vspace{1em}
\noindent\textit{All raw rows, calibration evidence, amendments, and scripts are
committed; every figure and table in this report is reproducible from
\texttt{results/} at the repository head.}

\end{document}
