\documentclass[sigconf,nonacm]{acmart}

\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{booktabs}
\emergencystretch=3em
\hbadness=10000
\sloppy

\begin{document}

\title{Architectures Differentiate --- and Where Their Premises Fail}
\subtitle{Status Report III --- the confirmatory local cell at $N{=}5$, a protocol-loss counterfactual, and swarm vote anatomy}

\author{Eric Medlock}
\affiliation{%
  \institution{University of North Carolina at Charlotte}
  \department{ITCS 6881 Independent Study}
  \city{Charlotte}
  \state{North Carolina}
  \country{USA}
}
\email{emedlock@charlotte.edu}

\begin{abstract}
The first confirmatory cell of the pre-registered architecture study is complete: the
frozen frontier tier (v2.1) $\times$ three architectures $\times$ $N{=}5$ on the local
environment --- 540 runs, every cell exactly 5/5, one config hash. Two validation
results anchor the report. First, the tier froze at monolithic domain aggregates of
42/58/67\% and \emph{reproduced} at 43/58/67\% --- the calibration methodology holds on
the production serving stack. Second, that stack mattered: a pre-sweep check caught the
v2 code domain drifting from 48\% (LM~Studio) to 35\% (Ollama), forcing a mechanical
re-source (v2.1, 58\%) and a new guardrail --- \emph{calibration bands are verified on
the serving stack that produces the study data}. The headline: swarm 61.1\% $>$
monolithic 56.1\% $>$ agentic 47.2\%, at 3.0$\times$ and 1.9$\times$ monolithic's token
cost respectively. Both multi-agent results, however, decompose into mechanism findings
rather than architecture wins. Agentic's deficit is partly \emph{protocol-compliance
loss}: the reasoning model buries its APPROVE verdict mid-prose, the framework's
first-line parser misses it, and the resulting ``false revision'' loops destroy
correct, already-approved answers. A pre-declared counterfactual re-parse of saved
traces recovers agentic math from 33\% to 42\% --- parity with monolithic --- while
code and multihop are untouched: the residual gap is real architecture cost. Swarm's
win dissolves under vote anatomy: across all 180 swarm rows the vote is never
genuinely contested (three 2--1 code splits from serving jitter aside); its multihop
edge comes from two tasks flipped deterministically by the peer prompt. At
temperature~0 the pre-registered swarm is a system-prompt ablation at triple cost, not
an ensemble --- motivating a proposed SWARM~2.0 arm (peers at $T{>}0$), its amendment
drafted for logging before any of its data exist. Scope has shifted: the confirmatory
environments are Apple Metal (M5~Max, M4) and university HPC (CUDA), whose sweep is
running as this report is written; the Azure cell is reframed from a GPU~VM to an
\emph{exploratory managed-inference arm} (Azure AI Foundry, serverless DeepSeek-R1)
to run in parallel with SWARM~2.0.
\end{abstract}

\maketitle

\section{Context and changes since Report II}
Design, hypotheses, and metrics remain frozen in \texttt{PRE\_REGISTRATION.md}
(2026-07-01); every deviation is amendment-logged. Since Report~II (2026-07-09):

\textbf{Serving stack pinned to Ollama; tier re-frozen as v2.1.} All remaining
environments serve via native Ollama (CUDA boxes and HPC have no LM~Studio path). A
production-stack calibration check --- monolithic $N{=}5$ over the frozen v2 code
domain --- measured \textbf{35\% on Ollama vs.\ 48\% in the original LM~Studio
calibration}. The drift is real (same 4 items solid, 7 floored) and out-of-band, so the
domain re-sourced mechanically under the domain-aggregate rule: v2.1 code is a
deterministic 6+6 mix of first-validating BigCodeBench-Hard and BigCodeBench items,
which calibrated to \textbf{58\% in band on Ollama} (7 items solved, 5 unsolved).
Math and multihop carry unchanged. Methodological consequence, now a guardrail:
difficulty is relative not only to the \emph{model} (Report~II) but to the
\emph{serving stack}; bands are verified on the production stack before any
multi-architecture sweep.

\textbf{Environment scope reshaped.} The pre-registered Azure cell is reframed
(amendment pending advisor confirmation): a rented GPU VM would duplicate the HPC
cell's serving stack on worse hardware, so the confirmatory environments are Apple
Metal (M5~Max primary; M4~Mini as a cross-hardware point) and university HPC (NVIDIA
A40, CUDA). Azure re-enters as an \emph{exploratory} managed-inference arm instead:
Azure AI Foundry's serverless full DeepSeek-R1 (the 14B distill is not
serverless-deployable), which cannot hold the pinned artifact --- precision, batching,
and decoding are not controllable --- and is therefore framed as ``self-hosted pinned
artifact vs.\ managed serving stack,'' measuring accuracy deltas, latency
distribution, determinism at temperature~0, and \$/token (verified pricing
\$1.35/\$5.40 per~M tokens; a full $N{=}5$ sweep estimates $\sim$\$27). A Shadow
cloud-PC (RTX~A4500) served as the CUDA rehearsal environment ($N{=}1$, 2026-07-12),
validating NVML telemetry and surfacing two harness fixes (the CUDA telemetry
dependency install; the reasoning-model \texttt{thinking}-field answer recovery)
before the confirmatory runs.

\section{Confirmatory local cell: 540 runs}
Frozen v2.1 tier (36 items) $\times$ 3 architectures $\times$ $N{=}5$ on the M5~Max
(48\,GB, Metal, Ollama). Integrity: all 108 cells exactly 5/5 trials, single config
hash, zero missing rows. Failure mix: 211 reasoning errors, 23 budget exhaustions,
10 code-harness (tool) errors.

\begin{table}[h]
\caption{Local confirmatory cell ($N{=}5$): accuracy by domain, and mean cost.}
\label{tab:n5}
\begin{tabular}{lrrrr}
\toprule
 & Math & Code & Multihop & \textbf{Overall}\\
\midrule
Monolithic & \textbf{43\%} & \textbf{58\%} & 67\% & 56.1\%\\
Agentic    & 33\% & 50\% & 58\% & 47.2\%\\
Swarm      & 42\% & \textbf{58\%} & \textbf{83\%} & \textbf{61.1\%}\\
\midrule
\multicolumn{5}{l}{\small Mean output tokens/run: 2755 / 5148 / 8177 (1$\times$ / 1.9$\times$ / 3.0$\times$)}\\
\multicolumn{5}{l}{\small Median latency/run: 48\,s / 88\,s / 140\,s at $\sim$43 tok/s (Metal)}\\
\bottomrule
\end{tabular}
\end{table}

\textbf{The calibration reproduced.} Monolithic domain aggregates landed at
43/58/67\% against calibration's 42/58/67\% --- a one-point deviation in one domain
across 180 fresh runs. The tier holds its band placement on the production stack; the
pre-declared selection rule measured what it claimed to measure.

\section{Mechanism finding 1: the false revision}
Agentic trails everywhere --- but instrumentation shows part of the deficit is not the
\emph{architecture} at all. The pinned verifier prompt demands the verdict on the first
line (``APPROVE or REVISE''); the backend parses it with
\texttt{startswith("APPROVE")}. DeepSeek-R1 explains first and verdicts later, so
verbal approvals routinely fail the parse; worse, any verifier turn that exhausts the
token budget returns chain-of-thought (the empty-content fallback), which
\emph{structurally} cannot begin with APPROVE. Each ``false revision'' forces another
Executor$\to$Verifier loop against an already-approved answer; the loop frequently
ends with an empty final answer. In the cleanest exhibit the executor answered 70
(gold: 70), the verifier wrote ``\ldots is indeed 70. APPROVE'', and all five trials
scored zero.

Per a pre-declared amendment (logged before any counterfactual number was computed), a
mechanical re-parse of the saved traces asks what each agentic row would have scored
under lenient verdict reading (last standalone APPROVE/REVISE token decides; the
approved candidate is graded by the pinned grader; primary rows untouched).

\begin{table}[h]
\caption{Agentic: pinned vs.\ counterfactual (local $N{=}5$).}
\label{tab:cf}
\begin{tabular}{lrrr}
\toprule
Domain & Pinned & Counterfactual & $\Delta$\\
\midrule
Math     & 33\% & 42\% & +8.3\,pts\\
Code     & 50\% & 50\% & 0\\
Multihop & 58\% & 58\% & 0\\
\textbf{Overall} & 47.2\% & 50.0\% & +2.8\,pts\\
\bottomrule
\end{tabular}
\end{table}

Five rows flip, all math, all wrong$\to$right; none flip down. The decomposition is
sharp: on math, protocol loss explains agentic's \emph{entire} deficit (counterfactual
42\% = parity with monolithic 43\% and swarm 42\%); on code and multihop the gap is
genuine architecture cost --- verifier judgment errors and revision-loop budget burn.
The general lesson: agentic is the only architecture whose \emph{control flow} runs
through model prose, and reasoning models are unreliable at strict output protocols;
frameworks that key control flow on exact tokens silently convert approvals into
destructive revision loops.

\section{Mechanism finding 2: swarm never votes}
Report~II's exploratory data hinted the swarm's majority vote might be inert at
temperature~0; the confirmatory cell settles it. Across all 180 swarm rows: no
genuinely contested votes (60/60 math and 60/60 multihop unanimous or single-key;
three 2--1 code splits from serving jitter; 28 math abstention events from correlated
budget exhaustion --- all three peers truncating together in 8 of the 10 affected
rows). Greedy decoding ignores the
per-peer seed offsets, so the peers are near-clones and the pre-registered
``independent samples'' premise fails mechanically.

The headline multihop win (83\% vs.\ 67\%) reduces to exactly two tasks, each flipped
0/5$\to$5/5 deterministically by the peer \emph{prompt}: one where the peer prompt
elicits a different reasoning path (monolithic answers the wrong hop), and one of
answer specificity (monolithic's under-specified league name fails the string grader;
peers emit the full name). At temperature 0 the pre-registered swarm is a
system-prompt ablation at 3$\times$ cost, not an ensemble.

Consequence: a \textbf{SWARM 2.0} arm is proposed, its amendment drafted for logging
\emph{before any of its data exist} --- identical topology, prompts, and tasks, but
peers decode at $T{>}0$ with per-peer seeds (temperature chosen by a small
pre-declared probe), restoring the premise the majority vote requires. Swarm-1.0 rows remain the valid measurement of the
pre-registered design; the paper reports them as a diagnosed negative result with the
vote-anatomy evidence above.

\section{Cross-hardware reproducibility}
Three environments have now run the frontier tier at the pinned 14B:

\begin{table}[h]
\caption{Cross-environment view (v2.1 except Shadow = v2; $N$ as shown).}
\label{tab:xenv}
\begin{tabular}{llrrr}
\toprule
Env (accel.) & $N$ & Mono & Agentic & Swarm\\
\midrule
M5 Max (Metal)   & 5 & 56\% & 47\% & 61\%\\
M4 Mini (Metal)  & 1 & 58\% & --- & 61\%\\
Shadow A4500 (CUDA, v2) & 1 & 44\% & 39\% & 50\%\\
\bottomrule
\end{tabular}
\end{table}

Aggregates reproduce across accelerators and boxes (the Shadow cell ran the earlier v2
code domain, depressing its absolute numbers); per-item results do not reproduce at
$N{=}1$ --- temperature 0 does not force identical outputs across Metal and CUDA.
Decode throughput is memory-bandwidth arithmetic: the base M4 ($\sim$120\,GB/s) holds
$\sim$10.4 tok/s on the 9\,GB Q4 model, the M5~Max $\sim$43 tok/s, the A4500
($\sim$640\,GB/s) $\sim$39 tok/s --- each near the ceiling its bandwidth predicts.

\section{Secondary metric: LLM-as-judge}
The different-family judge (Llama-3.2-3B) processed all 540 rows (535 parseable
verdicts): judge-correct 83 / 83 / 86\% and mean quality 3.63 / 3.67 / 3.81 (of 4)
for monolithic / agentic / swarm. The
3B judge is systematically lenient against the binary grader but preserves the
ordering; it serves as corroboration, not as a result.

\section{Threats to validity}
Single model family; two accelerator classes but one vendor stack (Ollama) by design
--- stack effects are now a documented, guarded variable rather than an assumption.
The agentic counterfactual is post-hoc (though rule-fixed before computation) and
reported alongside, never instead of, the pinned metric. Swarm-1.0's inertness means
the study so far bounds only \emph{prompt} effects of the swarm topology;
ensemble effects await SWARM~2.0. Cross-domain accuracy comparisons remain
descriptive (bands sit at 42/58/67\%). $N{=}5$ at temperature 0 measures serving
jitter, not sampling variance.

\section{Next steps}
(1)~HPC confirmatory sweep (A40, CUDA): running now, ETA within the day; data intake,
integrity checks, the same counterfactual re-parse, judge pass, and analysis follow.
(2)~SWARM~2.0: temperature probe, amendment, then the corrected cell on HPC ---
run \emph{in parallel with} (3)~the Azure AI Foundry exploratory arm (thin
OpenAI-compatible client, per-row cost accounting, $N{=}1$ cost pilot first).
(4)~Full paper draft from the assembled findings. (5)~Stretch (time permitting, in
priority order): a framework-as-agentic-layer A/B (LangChain/CrewAI vs.\ the
hand-rolled loop --- a direct extension of the false-revision finding), HPC scaling
legs (32B, Qwen3.6-35B), and a budget-sensitivity leg at max\_tokens 12288.

\vspace{1em}
\noindent\textit{All raw rows, traces, amendments, the counterfactual script
(\texttt{scripts/agentic\_counterfactual.py}), and judge outputs are committed; every
number in this report is reproducible from \texttt{results/} at the repository head.}

\end{document}
